---
title: "prcomp, princomp, cmdscale, svd"
output:
  html_document:
    df_print: paged
---

This doc is hosted here : [http://drpowell.github.io/utils/pca-mds.html](http://drpowell.github.io/utils/pca-mds.html)

[PCA](https://en.wikipedia.org/wiki/Principal_component_analysis) and [Classical MDS](https://en.wikipedia.org/wiki/Multidimensional_scaling#Classical_multidimensional_scaling) are closely related.  PCA works from the original coordinates,  MDS works from a distance matrix.  If the distance matrix uses Euclidean distance, then a classical MDS and PCA produce the same projection.

These R snippets show how to produce the same projection using PCA or MDS in R using `prcomp`, `princomp`, `cmdscale` or `svd`.

---

For this example, use the `mtcars` dataset.  This is what it looks like
```{r}
plot(mtcars)
```

### prcomp

Really should scale=T for this type of data, but not the point of this exercise.

```{r}
pr1 <- prcomp(mtcars)
plot(pr1$x, type='n')
text(pr1$x, labels=rownames(pr1$x), ce=0.6)
```

### princomp

```{r}
pr2 <- princomp(mtcars)
locs <- cbind(-pr2$scores[,1], pr2$scores[,2])          # Refections are arbitrary, swap one dimension to match previous plots
plot(locs, type='n')
text(locs, labels=rownames(pr2$scores), ce=0.6)
```

### MDS - cmdscale

Compute distance matrix from euclidean distances.

```{r}
d <- dist(mtcars)
mds <- cmdscale(d)
plot(mds, type='n')
text(mds, labels=rownames(mds), ce=0.6)
```


### MDS Using SVD

As implemented in [Degust](https://github.com/drpowell/degust/blob/master/degust-frontend/degust-src/js/mds-plot.vue#L86).

```{r}
m <- scale(as.matrix(mtcars), center=T, scale=F)   # Columns are centered
decomp <- svd(m %*% t(m))
proj <- decomp$v %*% sqrt(diag(decomp$d))
proj[,2] <- -proj[,2]         # Refections are arbitrary, swap one dimension to match previous plots
plot(proj, type='n')
text(proj, labels=rownames(mtcars), ce=0.6)
```